---
title: "West coast laby alpha diveresity 1"
author: "Lindsay Alma, Becca Maher"
date: "03/11/25"

---
## Script for calculating alpha diversity statistics and testing for significant differences with metadata variables

### 1 Setup the working space
#### Libraries
```{r libraries, message=FALSE}

library(phyloseq)
library(multcomp)
library(picante)
library(ggplot2 )
library(tidyverse)
```

#### Functions
```{r functions, include = FALSE}
# Arc-sine transformation
asinTransform <- function(p) { asin(sqrt(p)) }
# Estimate faith's phylogenetic diversity 
estimate_pd <- function(phylo){
  
  # Error if input is not of class phylo
  if(class(phylo) != "phyloseq"){
    stop("Input file is not of class 'phyloseq'.")
  }
  
  # Error if no class phy_tree
  if(!(.hasSlot(phylo, "phy_tree"))){
    stop("Could not find tree slot in phylo object.")
  }
  
  # Transpose if needed
  # Adapted from phyloseq/vegan import
  OTU <- phyloseq::otu_table(phylo)
  if (taxa_are_rows(OTU)) {
    OTU <- t(OTU)
  }
  
  # Get matrix version of OTU table
  otutable <- as(OTU, "matrix")
  
  # Get phylogenetic tree from pyloseq object
  tree <- phyloseq::phy_tree(phylo)
  
  # Print status message
  message("Calculating Faiths PD-index...")
  
  # If object is greater than 10mb, then print status message
  if(object.size(otutable) > 10000000){
    message("This is a large object, it may take awhile...")
  }
  
  # Calculate Faith's PD-index
  pdtable <- picante::pd(otutable, tree, include.root = F)
  
  # Return data frame of results
  return(pdtable)
}
# Calculate standard error
sderr <- function(x) {sd(x)/sqrt(length(x))}
#+++++++++++++++++++++++++
# Function to calculate the mean and the standard deviation
# for each group
#+++++++++++++++++++++++++
# data : a data frame
# varname : the name of a column containing the variable
#to be summariezed
# groupnames : vector of column names to be used as
# grouping variables
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sderr(x[[col]]), na.rm=TRUE)
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}
```

#### Load the data
For all alpha and beta diversity statistics, I will be using the rarefied sOTU table. 
```{r load}
library(vegan)
library(phyloseq)


setwd("C:/Users/Lindsay Alma/Dropbox/Eelgrass/West Coast TagSeq primers/FullDataset")

# Load the phyloseq object
ps <- readRDS("ps_18S_lulu.RDS")

ps <- subset_samples(ps, Person!="Olivia")
ps <- subset_samples(ps, Person!="CB")

# remove non-target taxa
ps = subset_taxa(ps, (domain!="Bacteria") | is.na(domain)) #161 Bacteria taxa
ps = subset_taxa(ps, (kingdom!="Chloroplastida") | is.na(kingdom)) #42 Chloroplastida taxa
ps = subset_taxa(ps, (kingdom!="Animalia" | is.na(kingdom))) # 30
ps = subset_taxa(ps, (kingdom!="Alveolata" | is.na(kingdom))) #9
ps = subset_taxa(ps, (kingdom!="Amoebozoa" | is.na(kingdom))) #3
ps = subset_taxa(ps, (kingdom!="Holozoa" | is.na(kingdom))) #10
ps = subset_taxa(ps, (kingdom!="Rhodophyceae" | is.na(kingdom))) #4
ps = subset_taxa(ps, (phylum!="Diatomea" | is.na(phylum))) #9
ps = subset_taxa(ps, (phylum!="Ochrophyta" | is.na(phylum))) #5
ps = subset_taxa(ps, (genus!="Aplanochytrium" | is.na(genus))) #46
ps = subset_taxa(ps, (domain!="Unassigned") | is.na(domain)) #161 Bacteria taxa 
ps
#ps <- subset_samples(ps, Lesion_Type == "Crescent" | Lesion_Type == "Not crescent")
ps
ps <- prune_samples(sample_sums(ps) >= 383, ps)
ps
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps

#sample_sums(ps)
```

### Calculating alpha diversity statistics
Initialize matrices to store alpha diversity estimates
```{r init}
nsamp = nsamples(ps)

observed <- matrix(nrow = nsamp)
row.names(observed) <- sample_names(ps)

shannon <- matrix(nrow =nsamp)
row.names(shannon) <- sample_names(ps)


```
Calculate statistics
```{r stats, message = FALSE}
# Calculate observed
obs <- as.numeric(as.matrix(estimate_richness(ps, measures = "Observed")))
observed[ ,] <- obs
colnames(observed) [1] <- "observed"


# Calculate shannon
shan <- as.numeric(as.matrix(estimate_richness(ps, measures = "Shannon")))
shannon[ ,] <- shan
colnames(shannon) [1] <- "shannon"


```
Combine our estimates for observed, simpson, and faith's PD into one dataframe
```{r compile}
alpha <- cbind(observed, shannon)

head(alpha)

# Add the sample metadata into this dataframe
s <- data.frame(sample_data(ps))
alphadiv <- cbind(alpha, s)
head(alphadiv)
alphadiv <- alphadiv[,-5]
head(alphadiv)
write.csv(alphadiv, "alphadiv_823.csv")
write.csv(alphadiv, "C:/Users/Lindsay Alma/Dropbox/Eelgrass/West Coast TagSeq primers/FullDataset/alphadiv_823.csv")

```
